### 2017-01-26 manuscript.rmd
changed title, removed unused calibrations, moved around and deleted some text

# The importance of phylogenetic methods when comparing functional genomic data across species

@@

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have, however, neglected to use methods that consider the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative gene expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other. The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals. We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.

@@

	# These clade dates are from the KMRR code, they got them from http://timetree.org/
	calibration_times = data.frame(
		age.min =
			c( 20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86 ), 
		clade = 
			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
					"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
					"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
					"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
					"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"
			),
		stringsAsFactors=FALSE
	)
```

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary following duplication and speciation events. We focused on two possible sources of bias - node depth and branch length.

@@

Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species (shared orthologs vary from pair to pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

We previously described some concerns with their interpretations of these results [@Hejnol:2016hw]. The main interpretation issues we identified with their interpretation is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to phyla or occurred along the many other branches in the tree. By the stated logic this study could just as easily be a test of unique expression differences between classes, orders, or families, since each of the sampled animals come from a different one of these groups. It is far more likely that changes between such distantly related taxa do not correspond to taxonomic ranks in any way.  Phyla were invented as taxonomic convenience to refer to animal groups of unknown relationship to each other, so they are a statement of historical phylogenetic uncertainty by humans rather than an intrinsic biological phenomenon to be explained.

@@

Here we examine the support for the inverse hourglass model itself. We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on. As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably. The medians are close in value, and the range of mid phase includes the entire range of the early phase. Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores. We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure XXLevin b).

We also found several problems with the statistical tests that were used to evaluate the inverse hourglass hypothesis. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears much stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b). Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] comes to a very conclusion - the greater variance at the midpoint of development is specific to a single lineage, the ctenophore.

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr]. This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa]. Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here. This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods [@Felsenstein:1985ua]. This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression.

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be. A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.

@@

Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are directly applicable to CFG data. There are also unique challenges that will need to be addressed at this new interdisciplinary interface, though. One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined. In CFG analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits. This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa]. In addition to methods development at this interface, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.

@@

@@

```{r pairwise_plot, echo=FALSE, cache=TRUE, message=F, warning=F}	
	pairwise_summary %>% 
	filter( ! is.na( D ) ) %>%
	ggplot( aes( x=distance, y=difference, color=D ) ) + 
		geom_point( alpha=0.1 ) +
		geom_smooth( method=lm, se=FALSE ) +
		xlim( 0, 2000 )

> Supplementary Figure XXPairwise | Pairwise comparisons

>>>>

# Multiple pairwise comparisons between species strongly mislead functional genomics studies

@@

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative gene expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other. The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals. We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.

@@

	# These clade dates are from the KMRR code, they got them from http://timetree.org/
	calibration_times = data.frame(
		age.min =
			c( 92, 167, 8, 296, 29, 162 ), 
		clade = 
			c( "Euarchontoglires", "Mammalia", "Homininae", 
					"Amniota", "Catarrhini", "Theria"
			),
		stringsAsFactors=FALSE
	)
```

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of evolution following duplication and speciation events. We focused on two possible sources of bias - node depth and branch length. We found no evidence that either affacted our results (Supplementary materials).

@@

Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these active areas of research and instead applied an *ad hoc* pairwise comparison of ortholog gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species. This list of genes was different from species pair to species pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

@@

We previously described some concerns with their interpretations of these results [@Hejnol:2016hw]. The main issue is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree. Here we examine the support for the inverse hourglass model itself. We focus on the matrix of pairwise comparisons that their KS tests and their Figure 4d are based on. As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably. The medians are close in value, and the range of mid phase includes the entire range of the early phase. Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores.

The first problems we identified were general stastistical issues that are not problems with pairwise comparisons *per se*. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case.

The next issue we found is a general problem with pairwise comparisons. Inspecting all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison). The inverse hourglass turns into a bottle (Figure XXLevin b).

This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - one evolutionary change impacts multiple comparisons. This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua]. While we do demonstrate it impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be. A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.

@@

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr]. This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa].

Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are already directly applicable to CFG data. There are interesting new challenges at this interdisciplinary interface that will need to be addressed to fully realize the potential of phylogenetic comparative functional genomics stidues. One such challenge is that most phylogenetic comparative analyses of covariance between traits have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined. In CFG analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits. This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa]. In addition to methods development, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.

@@

#### Investigation of potential ascertainment bias

@@
